Writing Trash Bin

### 2024-03-18 paper.qmd
updated some of the perceptual tasks section

### Information asymetry in evaluated plots
The erroneous notion that "uncertainty visualisations" are a class of visualisations that depict "uncertainty" (something that is undefined) has allowed authors to compare plots that depict wildly different information with little consideration of the value of this practice.

Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] (*Cite gap 12*). However, what makes two visualisations equal in "information" is not consistent and is regularly mentioned without it being clear exactly what this information is. This is easily illustrated by the wide array of comments made by visualisation authors when they explain why two representations are being compared in their evaluation study. @Ibrekk1987 displayed 6 PDFs in their experiment because "formally equivalent representations are often not psychologically equivalent", but do not explain why what makes a representation "formally equivalent" or why only one representation was presented for the 95% CI, box plot, and CDF. @Hofman2020 comments that "theoretically" the sampling distribution of the mean and the prediction interval of a new observation are equal "so long as one knows the sample size", but does not recognise the several assumptions and statistical knowledge that is required to compare the two. *Other examples* (*Cite Gap 13*). @Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed. It is clear this lacking definition of "information" in a visual is reminicent of our problems with the definition of "uncertainty".

@@

### Repeating perceptual task experiments
Given the previous section, one might consider problems in uncertainty visualisation to be fixed if we could restrict evaluation experiments to cases where the information in each graphic is almost identical. Unfortunately this still often results in uninteresting or

Once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is identical to the visualisation process of any other variable, or at least this is how it is treated in the uncertainty visualisation experiments.

Computational equivalency in visualisation is also something that has been well researched. For simple tasks such as value extraction, there is a hierarchy to perceptual tasks where extracting visual information in some forms is easier than others. The hierarchy established 40 years ago by @Cleveland1984 is:

1) position along a common scale. 
2) position along a non-aligned scale. 
3) length. 
4) angle/slope. 
5) area. 
6) volume. 
7) colour

> Edit notes: need to add the notes from the susan lit review that has the more recent perceptual task list

This hierarchy has had additions and adjustments over the years. For example, this hierarchy is a general rule, but the hierarchy can change from individual to individual. Additionally, there are other graphical rules to consider such as guestalt heuristics, broader methods of extraction, and attention principles. These established visualisation concepts allow us to anticipate the ease with which certain pieces of information will be extracted from a plot.

A bizarre feature of uncertainty visualisation literature is that it does not work to build upon these existing principle or identify the ways in which uncertainty visualisations may diverge from these rules, but rather it seems to exist with complete indifference to the research in information visualisation at large. The most common uncertainty visualisation evaluation study asks participants to perform some value abstraction either explicitly by asking participants to report values or vaguely by asking participants to find regions of least certainty, rank plots, report their confidence, and/or make decisions according to uncertainty information. Experimenters then compare these results based on the participants accuracy, preference, confidence, and/or speed @Hullman2016. Most commonly, these value abstraction papers compare two different visualisations of uncertainty on a set of these questions, however, since there is no rules for what is or is not an uncertainty visualisation the graphics that are compared often differ in both computation *and* information. @Hullman2018 asks participants questions that would require a prediction interval but provided a sampling distribution and argued that the two could be considered equivalent so long as participants understood the difference between a confidence interval and a prediction interval, knew the sample size, and knew the formula to convert the confidence interval to a prediction interval. The authors themselves note that this is rediculous, which begs the question. Uncertainty visualisaiton papers only seem to require that the plots depict "uncertainty" which has already been established to be poorly defined. The complete lack of formality is easy to see if we actually plot the visualisations compared across evaluation experiments according to the grammar of graphics *in the figure below*:

> Edit note: need to include a graphic that illustrates the wide spread of mathematical information and visual techniques that are visualised in these experiments.

Reading the uncertainty visalisation literature will make the importance of information and computation incredibly salient. If plots that have different information are compared, whichever directly depicts the ground truth statistic (or has information that can easily be convered to that ground truth statistic) will perform the best *cite*. If plots that have the same information are compared, the plot that requires a perceptual task that is higher on the hierarchy will perform the best *cite*. Therefore, when it comes to any task that involves value abstraction, many uncertainty visualisation evaluation studies are simple repeating already established results in visualisation.

Several uncertainty visualisation literature reviews have commented on the immense difficulty of combining results across experiments, but none have been able to pinpoint the reason. @Kinkeldey2014 suggested a task based taxonomy that organised the plots by the question or purpose of the visualisation, and even suggested that communication tasks (such as value retrival of uncertainty values) should just follow traditional cartography rules, @Spiegelhalter2017 simply ignored the existence of uncertainty visualisation as a subfield and just suggested we follow established rules within uncertainty visualisation, @Hullman2016 said that commented on the large amount of noise generated from uncertainty visualisation evaluation methods that led to a difficulty in combining results, @Griethe2006 did not meaningfully combine the work and instead listed findings in the field, but also did not comment on the lack of generalisable results. While we largely agree with the difficulties of these authors, we do not agree that the solution is a task based taxonomy or better questions. While we agree the lack of attention to tasks and appropriate questions has generated problems in the field, there is a much more pressing issue, which is that uncertainty visualisation authors don't seem to understand *why* uncertainty visualisation different to typical visualisations.

The class of "uncertainty visualisation" does not seem to have any rules for what belongs in the group and what makes it different to a "typical" visualisation is rarely discussed. *removed bit about two motivations* Neither of these motivations are remotely relevant when we are evaluating the simple task of value extraction. The psychological effects of uncertainty are only relevant for decision making and have nothing to do with value abstraction, and the information overload issue is not specific to uncertainty and is simply an example of a larger dimensionality issue.
Therefore, these explanations for the difference fall short of explaining *why* uncertainty should be treated any differently than any other variable.

There is also the common view that uncertainty as a dimensionality problem. This perspective is not quite right and the literature seems to be about 50/50 on whether uncertainty is a psychological or dimensionality problem.  It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen". The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional]. Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all. The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant". Including uncertainty is worth very little if no attention is left to see it. This does not mean we cannot put a *large* amount of information in a graphic. Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important. The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey. No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.

This issue has come up in many different forms in the uncertainty visualisation literature. Typically as a "task dependence" on the performance of uncertainty visualisations. It is repeatedly identified as a problem in previous reviews of the uncertainty visualisation literature [@Kinkeldey2014; @Hullman2016] as well as across many sub domains and applications [@Wallsten1997; @Fischhoff2014; @Meng2021; @Amar2005]. The fact that this conclusion is repeatedly reached shows both the importance and the lack of acknowledgement this concept receives. @Fischhoff2014  takes this approach by considerting how decisions with different levels of flexibility should have different uncertainty visualisations. If the user has no choice they should simply be made aware of the risk, if they have several options, the user needs to see the uncertainty in each, if they need to come up with the options themselves, they need to see how things work and how the system fails. @Kinkeldey2014 suggested that the nature of the task plays an important role and believed the task dependence for uncertainty visualisation could explain some of the inconsistency in outcomes.

@@

## Fixes within "uncertainty visualisation"
Attempts to rectify these conflicts always flutter about the surface and do not seem to grasp the core of this problem. For example, @wu2023rational noticed a series of papers with poorly defined tasks and suggested comparing study participants to a "rational agent" to help authors understand and remedy this issue. @Kim2019 believed an understanding of participants prior beliefs could assist in interpreting uncertainty visualisation results that tend to be incredibly noisy. What these suggestions fail to understand is that the entire practice of identifying "true" facts about "unertainty visualisation" is futile. It is impossible to make a statement about the properties of an "uncertainty visualisation" in the same way it is impossible to comment on $\frac 0 0 = 1$. Division by zero is undefined, so the statement is ill formed, similarly, "uncertainty" is not defined, so statements *about* "uncertainty" including statements about "uncertainty visualisation" are ill formed. Ill-formed statements can neither be true nor false and cannot be part of a logical argument, so experiments and research that build upon this ill-formed concept are neither true nor false, they simply exist. Someone who is less familiar with the rules of logical argument may believe this understanding of the problem to an insufferable nitpick that creates impossible to live up to standards, however the consequences of this definitional issue are *very* real and costly.  The studies that investigate "uncertainty visualisation" may be identifying *something*, however there is good reason to believe that *something* is neither interesting nor does can it contribute to a larger theory of visualisation. This is something we illustrate in the following this section.

# Uncertainty as noise and Uncertainty as signal

@@

>>>>

### Information asymetry in evaluated plots
Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] (*Cite gap 12*). However, what makes two visualisations equal in "information" is not consistent and is regularly mentioned without it being clear exactly what this information is. This is easily illustrated by the wide array of comments made by visualisation authors when they explain why two representations are being compared in their evaluation study. @Ibrekk1987 displayed 6 PDFs in their experiment because "formally equivalent representations are often not psychologically equivalent", but do not explain why what makes a representation "formally equivalent" or why only one representation was presented for the 95% CI, box plot, and CDF. @Hofman2020 comments that "theoretically" the sampling distribution of the mean and the prediction interval of a new observation are equal "so long as one knows the sample size", but does not recognise the several assumptions and statistical knowledge that is required to compare the two. *Other examples* (*Cite Gap 13*). @Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed. It is clear this lacking definition of "information" in a visual is reminicent of our problems with the definition of "uncertainty".

@@

### Repeating perceptual task experiments
Given the previous section, one might consider problems in uncertainty visualisation to be fixed if we could restrict evaluation experiments to cases where the information in each graphic is almost identical. Unfortunately this still often results in uninteresting or obvious results.

If two graphics are visually different but identical in the information they contain, they must differ in how that information is depicted. 
Once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is theoretically identical to the visualisation process of any other variable. For simple tasks such as value extraction, there is a hierarchy to perceptual tasks where extracting visual information in some forms is easier than others. The hierarchy was originally established 40 years ago by @Cleveland1984, below is an updated version summarised by @Vanderplas2020:

1) Position along a common scale. 
2) Position along a non-aligned scale. 
3) Length, direction, angle, slope
4) Area
5) Volume, density, curvature
6) Shading, color saturation, color hue
7) Discriminable shape
8) Indiscriminable shape

This hierarchy is a good general rule, however it can change from person to person [@Davis2022] Additionally, there are other graphical rules to consider such as guestalt heuristics, broader methods of extraction, and attention principles [@Vanderplas2020]. These established visualisation concepts allow us to anticipate the ease with which certain pieces of information will be extracted from a plot. We can use these concepts to understand the computational complexity of a graphic. A bizarre feature of the uncertainty visualisation literature is that it does not work to build upon these existing principles or identify the ways in which uncertainty visualisations may diverge from these rules. These building block concepts of visualisation are seldom mentioned.

It is difficult to find examples of uncertainty visualisation experiments where the plots do contain the same information, however when they do, the results align with existing information visualisation research.

*Other examples* (*Cite Gap 15*).

If plots that have different information are compared, whichever directly depicts the ground truth statistic (or has information that can easily be convered to that ground truth statistic) will perform the best. If plots that have the same information are compared, the plot that requires a perceptual task that is higher on the hierarchy will perform the best. Therefore, when it comes to any task that involves value abstraction, many uncertainty visualisation evaluation studies are simple repeating already established results in statistics or visualisation.

Therefore, these explanations for the difference fall short of explaining *why* uncertainty should be treated any differently than any other variable.  The psychological effects of uncertainty are only relevant for decision making and have nothing to do with value abstraction, and the information overload issue is not specific to uncertainty and is simply an example of a larger dimensionality issue.

@@

# Uncertainty as noise and Uncertainty as signal
In this section we will discuss how *most* attempts to measure the effects of uncertainty are either value extraction experiments with an extra step, or are measuring uncertainty using a proxy that does not actually represent uncertainty. In this section we will talk about the issues of trying to measure uncertainty, the real purpose of uncertainty visualisation, and suggestions in how we can better measure uncertainty in our visualisation experiments.

## Most experiments are value extraction experiments with extra steps
Those with a keen eye might have read the previous sections and noticed our statements were only relevant to evaluation experiments where participants must extract a particular value from a plot. Of course, these experiments are not reflective of "real world" uses of uncertainty visualisations, we do not want to visualise uncertainty so that we can answer questions such as *what is the probability of even A*. In reality uncertainty is useful for decision making and eliciting trust from an audience. This section will explain why, despite appearing to be different, these experiments are actually still value extraction experiments and thus have the same pitfalls.

### Types of Uncertainty Visualisation Papers

@@

The uncertainty visualisation evaluation study asks participants to perform some value abstraction either explicitly by asking participants to report values or vaguely by asking participants to find regions of least certainty, rank plots, report their confidence, and/or make decisions according to uncertainty information [@Hullman201]. Experimenters then compare these results based on the participants accuracy, preference, confidence, and/or speed [@Hullman2016].

### 2024-03-18 writingdumpster.qmd
finished information asymetry examples
updated some of the perceptual tasks section

# Extra Section Cite Gaps
Cite Gap - New Section: Logical fallacy interpretation of mathematically unanswerable questions. If I tell you my cousins age, and then ask you to guess my age, it would not be unreasonable to guess the age I provided for my cousin. Citing this work to state "people incorrectly believe my cousins age is my age" would
- Deterministic questions about a random event cannot be answered. 
- @uncertchap2022 took the results of other paper to mean "when participants viewed mean temperature forecasts that included 95% confidence intervals depicted as bars with end caps, they incorrectly believed that the error bars represented high and low temperatures" (this one will depend on that study)

@@

@@

# Main section outtakes
- The process of working out whether or not a visualisation shows a piece of information, and if it does, what perceptual task it is mapped to and what assumptions or heuristics need to be used, is *incredibly* time consuming. While this method does predict the results for these experiments with remarkable accuracy, to do it for a large number of plots, especially when the ground truth statistics and visualisation are incredibly complicated is just not feasible for a reviewer to do.
- This is likely because The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics. The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
- We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline. We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence. Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics would we?
- What exactly is uncertainty, then? Let us consider the concepts related to uncertainty that have been discussed so far. We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline. We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence. Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics. For example, I wouldn't describe the population parameters of a distribution I have simulated to be "uncertain"

>>>>

Cite Gap 12: Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared

Cite Gap 13: Examples of what is "information" in a visualisation

Cite Gap 14: Examples of mathematically obvious results

Cite Gap 15: Examples of replicated perceptual task experiments

## Extra Section Cite Gaps
Cite Gap - New Section: Logical fallacy interpretation of mathematically unanswerable questions. If I tell you my cousins age, and then ask you to guess my age, it would not be unreasonable to guess the age I provided for my cousin. Citing this work to state "people incorrectly believe my cousins age is my age" would
- Deterministic questions about a random event cannot be answered. 
- @uncertchap2022 took the results of other paper to mean "when participants viewed mean temperature forecasts that included 95% confidence intervals depicted as bars with end caps, they incorrectly believed that the error bars represented high and low temperatures" (this one will depend on that study)

@@

@Kim2019 believed an understanding of participants prior beliefs could assist in interpreting uncertainty visualisation results that tend to be incredibly noisy.

@@

# Main section outtakes
There is also the common view that uncertainty as a dimensionality problem. This perspective is not quite right and the literature seems to be about 50/50 on whether uncertainty is a psychological or dimensionality problem.  It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen". The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional]. Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all. The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant". Including uncertainty is worth very little if no attention is left to see it. This does not mean we cannot put a *large* amount of information in a graphic. Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important. The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey. No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.

This issue has come up in many different forms in the uncertainty visualisation literature. Typically as a "task dependence" on the performance of uncertainty visualisations. It is repeatedly identified as a problem in previous reviews of the uncertainty visualisation literature [@Kinkeldey2014; @Hullman2016] as well as across many sub domains and applications [@Wallsten1997; @Fischhoff2014; @Meng2021; @Amar2005]. The fact that this conclusion is repeatedly reached shows both the importance and the lack of acknowledgement this concept receives. @Fischhoff2014  takes this approach by considerting how decisions with different levels of flexibility should have different uncertainty visualisations. If the user has no choice they should simply be made aware of the risk, if they have several options, the user needs to see the uncertainty in each, if they need to come up with the options themselves, they need to see how things work and how the system fails. @Kinkeldey2014 suggested that the nature of the task plays an important role and believed the task dependence for uncertainty visualisation could explain some of the inconsistency in outcomes.

Several uncertainty visualisation literature reviews have commented on the immense difficulty of combining results across experiments, but none have been able to pinpoint the reason. @Kinkeldey2014 suggested a task based taxonomy that organised the plots by the question or purpose of the visualisation, and even suggested that communication tasks (such as value retrival of uncertainty values) should just follow traditional cartography rules, @Spiegelhalter2017 simply ignored the existence of uncertainty visualisation as a subfield and just suggested we follow established rules within uncertainty visualisation, @Hullman2016 said that commented on the large amount of noise generated from uncertainty visualisation evaluation methods that led to a difficulty in combining results, @Griethe2006 did not meaningfully combine the work and instead listed findings in the field, but also did not comment on the lack of generalisable results. While we largely agree with the difficulties of these authors, we do not agree that the solution is a task based taxonomy or better questions. While we agree the lack of attention to tasks and appropriate questions has generated problems in the field, there is a much more pressing issue, which is that uncertainty visualisation authors don't seem to understand *why* uncertainty visualisation different to typical visualisations.

Attempts to rectify these conflicts always flutter about the surface and do not seem to grasp the core of this problem.  What these suggestions fail to understand is that the entire practice of identifying "true" facts about "unertainty visualisation" is futile. It is impossible to make a statement about the properties of an "uncertainty visualisation" in the same way it is impossible to comment on $\frac 0 0 = 1$. Division by zero is undefined, so the statement is ill formed, similarly, "uncertainty" is not defined, so statements *about* "uncertainty" including statements about "uncertainty visualisation" are ill formed. Ill-formed statements can neither be true nor false and cannot be part of a logical argument, so experiments and research that build upon this ill-formed concept are neither true nor false, they simply exist. Someone who is less familiar with the rules of logical argument may believe this understanding of the problem to an insufferable nitpick that creates impossible to live up to standards, however the consequences of this definitional issue are *very* real and costly.  The studies that investigate "uncertainty visualisation" may be identifying *something*, however there is good reason to believe that *something* is neither interesting nor does can it contribute to a larger theory of visualisation. This is something we illustrate in the following this section.

- The process of working out whether or not a visualisation shows a piece of information, and if it does, what perceptual task it is mapped to and what assumptions or heuristics need to be used, is *incredibly* time consuming. While this method does predict the results for these experiments with remarkable accuracy, to do it for a large number of plots, especially when the ground truth statistics and visualisation are incredibly complicated is just not feasible for a reviewer to do.
- This is likely because The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics. The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
- We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline. We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence. Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics would we?
- What exactly is uncertainty, then? Let us consider the concepts related to uncertainty that have been discussed so far. We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline. We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence. Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics. For example, I wouldn't describe the population parameters of a distribution I have simulated to be "uncertain"
